---
title: "Human Performance Final Report"
author: "Jared Smith"
date: "2023-05-07"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggdark)
library(plotly)
library(ggpubr)
library(xgboost)
library(MASS)
library(caret)
library(corrplot)
library(ggExtra) # Load ggExtra
library(xgboost) # Load XGBoost
source("a_insights_shap_functions.r") # Load SHAP functions
library(Metrics) # Load metrics
library(pROC) # Load proc 
day <- read_csv("day_approach_maskedID_timeseries.csv")
week <- read_csv("week_approach_maskedID_timeseries.csv")
# Remove rows with missing or invalid values in injury column
week <- week[complete.cases(week$injury), ]
```
```{r}
summary(week)
```


```{r}

noninjured <-week[week$injury == 0,]
summary(noninjured)
injured <- week[week$injury == 1, ]
summary(injured)
injured$injury = (as.factor(injured$injury))
levels(injured$injury) <- c("Injured")

noninjured$injury = (as.factor(noninjured$injury))
levels(noninjured$injury) <- c("NotInjured")

```


```{r}
injury_dist <- week %>%
  group_by(`Athlete ID`, injury) %>%
  summarise(count = n()) %>%
  spread(injury, count, fill = 0) 
injury_dist
colnames(injury_dist) <- c("Athlete ID", "Not Injured", "Injured")

#p <- ggplot(injury_dist, aes(x=`Athlete ID`, fill=Injured))+ # Set dataset and aesthetics
#  geom_density(alpha = 0.3, fill='red3') 
#p
```
```{r}
injury_sum <- week %>%
  subset(week$injury==1,) 
injury_sum <- injury_sum %>%
  group_by(`Athlete ID`, injury) %>%
  mutate(Freq = n()) 
injury_sum

ggplot(injury_sum, aes(x=`Athlete ID`, fill=Freq))+
  geom_bar()
```


```{r}
# Create plot
p <- ggplot(injury_sum, aes(x = `max exertion`, fill = injury)) + # Set dataset and aesthetics
  geom_density(alpha = 0.3, fill='red3') + 
  labs(x = "Max Exertion", y = "Injuries", # Set labels for plot 
       title = "Injuries by Max Exertion") +
  theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) + # Remove grid 
  dark_theme_bw() # Turn theme to dark mode

p1 <- ggplot(injury_sum, aes(x = `avg recovery`, fill = injury)) + # Set dataset and aesthetics
  geom_density(alpha = 0.3, fill ="blue2") + 
  labs(x = "Average Recovery", y = "Injuries", # Set labels for plot 
       title = "Injuries by Average Recovery") +
  theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) + # Remove grid 
  dark_theme_bw() # Turn theme to dark mode

p <- ggarrange(p, p1, ncol = 2, widths = c(3,3), common.legend = TRUE, legend = "bottom")
print(p)
```

For Weeks that an Athlete was Injured this was their average exertion
```{r}
# Create a ggplot boxplot
p <- ggplot(week, aes(x = injury, y = `avg exertion`, fill=injury)) +
  geom_boxplot() 
p
# Create an interactive plotly version of the ggplot
ggplotly(p) %>%
  layout(
    title = "Boxplot of Average Exertion by Injury Status",
    xaxis = list(title = "Injury Status"),
    yaxis = list(title = "Average Exertion")
  )
```


```{r}
p <- week %>%
  ggplot(aes(x = Date)) +
  geom_density(aes( fill = injury), alpha = 0.3) +
  scale_fill_manual(values = c("red2", "green")) +
  labs(title = "Bar Chart and Density Plot of Average Exertion by Injury Status",
       x = "Date", y = "Density", fill = "Injury Status")
p
```


```{r}
summary(week)
# distance sum over 3 weeks 
# week 0-2 (Distance in week 0 and the average across 1 and 2)
```




```{r}
ac_7_21 <- week$`total kms`/((week$`total kms.1` + week$`total kms.2`)/2)

ac_7_21[is.infinite(ac_7_21)] <- NA
```


```{r}
week$ac_7_21 <- ac_7_21
p <- week %>%
  ggplot(aes(x = ac_7_21)) +
  geom_density(aes( fill = factor(injury)), alpha = 0.3) +
  scale_fill_manual(values = c("red2", "green")) +
  labs(title = "Bar Chart and Density Plot of Average Exertion by Injury Status",
       x = "Date", y = "Density", fill = "Injury Status") +
  xlim(0,3)
p
```


## Derived Variables 

```{r}
ae_7_21 <- week$`avg exertion`/((week$`avg exertion.1` + week$`avg exertion.2`)/2)
ae_7_21[is.infinite(ae_7_21)] <- NA


exer_v_rec_7_21 <- week$`max exertion`/((week$`avg recovery.1`+ week$`avg recovery.2`)/2) ## Exertion 0 / recovery 1 over 3 weeks 
exer_v_rec_7_21[is.infinite(exer_v_rec_7_21)] <- NA

er_7_21 <- week$`avg exertion`/((week$`avg recovery.1` + week$`avg recovery.2`)/2)
er_7_21[is.infinite(er_7_21)] <- NA

rec_maxexer_7_21<- week$`avg recovery`/((week$`max exertion.1` + week$`max exertion.2`)/2)
rec_maxexer_7_21[is.infinite(rec_maxexer_7_21)] <- NA




week$ae_7_21 <- ae_7_21
threshold <- 2.25
week$ae_7_21 <- ifelse(week$ae_7_21 > threshold, mean(week$ae_7_21), week$ae_7_21)
summary(week$ae_7_21)

week$exer_v_rec_7_21 <- exer_v_rec_7_21 
threshold <- 2.25
week$exer_v_rec_7_21 <- ifelse(week$exer_v_rec_7_21 > threshold, mean(week$exer_v_rec_7_21), week$exer_v_rec_7_21)
summary(week$exer_v_rec_7_21)

week$er_7_21 <- er_7_21
threshold <- 2.25
week$er_7_21 <- ifelse(week$er_7_21 > threshold, mean(week$er_7_21), week$er_7_21)
summary(week$er_7_21)

week$rec_maxexer_7_21 <- rec_maxexer_7_21
threshold <- 2.25
week$rec_maxexer_7_21 <- ifelse(week$rec_maxexer_7_21 > threshold, mean(week$rec_maxexer_7_21), week$rec_maxexer_7_21)
summary(week$rec_maxexer_7_21)


```


```{r}
model <- week[,c(1:3, 12:14, 16, 17, 20, 68, 73:76)]
model <- na.omit(model)

full_model <- glm(injury ~ ., data = model, family = binomial())
summary(full_model)
step_model <- stepAIC(full_model, direction = "forward", scope = list(lower = ~1, upper = ~.), trace = FALSE)
summary(step_model)
```


### New Analysis 
## Lets investigate injuries as a part of our data set 
Looks as if our current percentage of injuries as opposed to training days is 1.42 %
```{r}
training_days <- sum(week$`nr. sessions` > 0)
sum(as.numeric(week$injury)) / training_days
summary(as.factor(week$injury))
```

```{r}
athlete <- week %>%
  group_by(`Athlete ID`) %>%
  summarise(Injuries = sum(as.numeric(injury))) %>%
  arrange(desc(Injuries)) %>%
  head(10)
athlete

```

```{r}
Injury_Plot <- ggplot(athlete, mapping = aes(x = Injuries)) +
  geom_density(alpha=0.5, fill="maroon") +
  labs(x = "Total Injuries",  
       title = "Distribution of Injuries among Athletes")
Injury_Plot
```

## Lets focus on the most injured athlete
```{r}
athlete1 <- week[week$`Athlete ID` == "26",]

# Create plot
Athlete_Workloads <- ggplot(athlete1, # Set data
              aes(x = `Date`, y = `avg exertion`, color = factor(injury))) + # Set aesthetics
  geom_point(alpha = 0.5) + # Set geom_point for scatter plot
  labs(x = "Days",  # Set labels
       title = "Athlete 26 Injuries",
       fill = "Injuries") +
  theme_bw() +
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) +  # Remove grid 
  scale_color_manual(values = c("0" = "blue", "1" = "red"), # Set color values
                     labels = c("0" = "Healthy", "1" = "Injured"))

# Generate graph
ggMarginal(Athlete_Workloads , groupFill = TRUE,
           type = "histogram")
```

```{r}
pivotdat <- pivot_longer(athlete1[, c("min recovery", "avg recovery", "max recovery", "min exertion", "avg exertion", "max exertion", "min training success", "avg training success", "max training success", "Athlete ID", "injury")],
                       !c("Athlete ID", "injury")) # Set variables to use as ID
pivotdatall <- pivot_longer(week[, c("min recovery", "avg recovery", "max recovery", "min exertion", "avg exertion", "max exertion", "min training success", "avg training success", "max training success", "Athlete ID", "injury")],
                       !c("Athlete ID", "injury")) # Set variables to use as ID
```

```{r}
g <- ggplot(pivotdat,
              aes(x = value, fill = factor(injury))) + # Set aesthetics
  geom_density(alpha = 0.3) + # Set geom density for density plot
  labs(x = "Percieved Metric Values",  # Set labels
       title = "Injuries v Percieved Metric Values Leading Up to Injury",
       fill = "injury") +
  facet_wrap(~name, scale = "free") +
  theme_bw() + # Set theme
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) + # Remove grid 
   scale_fill_manual(values = c("0" = "blue", "1" = "red"), # Set color values
                     labels = c("0" = "Healthy", "1" = "Injury"))

# Generate graph
g

gall <- ggplot(pivotdatall,
              aes(x = value, fill = factor(injury))) + # Set aesthetics
  geom_density(alpha = 0.3) + # Set geom density for density plot
  labs(x = "Percieved Metric Values",  # Set labels
       title = "Injuries v Percieved Metric Values Leading Up to Injury",
       fill = "injury") +
  facet_wrap(~name, scale = "free") +
  theme_bw() + # Set theme
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) + # Remove grid 
   scale_fill_manual(values = c("0" = "blue", "1" = "red"), # Set color values
                     labels = c("0" = "Healthy", "1" = "Injury"))

# Generate graph
gall
```

## Athlete ID 26 Week before injury
```{r}
# Lets select top 10 most injured athletes to see if there are any outstanding similarities
athletes <- week[week$`Athlete ID` %in% athlete$`Athlete ID`,]
# Create graph
athlete_ex <- ggplot(athlete1, # Set data
              aes(x = `avg recovery.1`, y = `max exertion` , color = factor(injury))) + # Set aesthetics
  geom_point(alpha = 0.3) + # Set geom point for scatter plot
  labs(x = "Avg Recovery week before Injury",  # Set labels
       y = "Max Exertion week of Injury",
       title = "Max Exertion versus Average Recovery before injury",
       fill = "injury") +
  theme_bw() + # Set theme
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) + # Remove grid 
   scale_color_manual(values = c("0" = "blue", "1" = "red"), # Set color values
                     labels = c("0" = "Healthy", "1" = "Injury"))

# Generate graph
ggMarginal(athlete_ex, groupFill = TRUE)
```

```{r}

# Create graph
week_ex <- ggplot(week, # Set data
              aes(x = `avg recovery.1`, y = `max exertion`,color = factor(injury))) + # Set aesthetics
  geom_point(alpha = 0.3) + # Set geom point for scatter plot
  labs(x = "Avg Recovery week before Injury",  # Set labels
       y = "Max Exertion week of Injury",
       title = "Max Exertion versus Average Recovery before injury",
       fill = "injury") +
  theme_bw() + # Set theme
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank()) + # Remove grid 
   scale_color_manual(values = c("0" = "blue", "1" = "red"), # Set color values
                     labels = c("0" = "Healthy", "1" = "Injury"))
# Generate graph
ggMarginal(week_ex, groupFill = TRUE)
```


```{r}
set.seed(42069) # Set seed for reproducibility
trainIndex <- createDataPartition(y = week$injury, p = 0.75, list = FALSE)
trainData <- week[trainIndex, ]
testData <- week[-trainIndex, ]
validData <- week[week$injury==1,]
# c(1:4,6, 12:22, 73:76)
# Create training data
dtrain_1 <- xgb.DMatrix(data = as.matrix(trainData[,c(1:4,6, 12:22, 36:44, 58:66, 74:77)]), label = ((trainData$injury)))
# Create test data
dtest_1 <- xgb.DMatrix(data = as.matrix(testData[,c(1:4,6, 12:22, 36:44, 58:66, 74:77)]), label = ((testData$injury)))
```

```{r}
fit_1 <- xgboost(dtrain_1,  # Set data set to use
                 nrounds = 200, # Set number of rounds
               eta = 0.05 , 
               verbose = 1, # 1 - Prints out fit
                print_every_n = 20, # Prints out result every 20th iteration
               
               objective = "binary:logistic", # Set objective
               eval_metric = "auc",
               eval_metric = "error")
```

```{r}
preds_injury <- predict(fit_1, dtest_1)
roc1 = roc((testData$injury), preds_injury)
plot.roc(roc1, print.auc = TRUE, col = "red", print.auc.col = "red")
```


```{r}
pred_class <- rep(0, length(preds_injury))

# If predicted probability above cutoff set as 1
pred_class[which(preds_injury > 0.0145)] <- 1
```

```{r echo=FALSE}

confusionMatrix(data = as.factor(pred_class), # Set predicted values
                reference =as.factor(testData$injury)) # Set true values
```

```{r}  
shap_result_1 <- shap.score.rank(xgb_model = fit_1, 
                X_train = as.matrix(trainData[,c(1:4,6, 12:22, 36:44, 58:66, 74:77)]),
                shap_approx = F)
```

```{r}
shap_long_1 = shap.prep(shap = shap_result_1,
                           X_train =  as.matrix(trainData[,c(1:4,6, 12:22, 36:44, 58:66, 74:77)]), 
                           top_n = 8)
```

```{r}
plot.shap.summary(data_long = shap_long_1)
```


## Future Steps Could possibly redo study to focus on the injuries of the top 10 most injured runners to see if they have any signficant metrics that lead to injury. 

```{r}
athletes <- week[week$`Athlete ID` %in% athlete$`Athlete ID`,]
```

